---
title: "Project 05 — Clinical Trial Simulation: Sample Size for EC₅₀ Precision"
subtitle: "1000 replicates · rxode2 + future.apply · Power curve and operating characteristics"
date: today
---

```{r setup, include=FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(gt)
source(here("R", "helpers.R"))
theme_set(theme_bw())
```

> **Simulated data.** Trial replicates are generated from the PK/PD model
> estimated in Project 02. True parameters are carried over unchanged.

## Background

Clinical trial simulation (CTS) quantifies the operating characteristics of
a planned study design before committing resources. For dose-ranging studies
the key question is not hypothesis testing but parameter estimation precision:
can the EC₅₀ be estimated with enough accuracy to support dose selection in a
subsequent confirmatory trial?

## Question

For a Phase 2 dose-ranging study (4 doses, placebo-controlled), what sample
size per arm achieves ≥ 80% power to estimate EC₅₀ within ± 30% of the true
value? What are the type I error rate and estimation bias?

## Study Design

| Design element | Value |
|----------------|-------|
| Doses | 50, 100, 200, 400 mg oral QD |
| Arms | Placebo + 4 active doses |
| n per arm | 20, 40, 60, 80, 120 (5 scenarios) |
| Endpoint | Continuous biomarker at 24 h |
| Analysis | Non-linear least squares Emax fit to simulated arm means |
| Success criterion | 95% CI for EC₅₀ contains [0.7×, 1.3×] true EC₅₀ |
| Replicates | 1000 per scenario |

## Model

PK/PD model from Project 02 (1-cmt oral PK, sigmoid Emax PD), re-used via rxode2.
IIV on CL, V, E₀, E~max~, EC₅₀. RUV proportional (PK) and additive (PD).

## Results

```{r power-curve, fig.width=9, fig.height=5, fig.cap="Power (% replicates with EC50 estimate within ±30% of true value) vs sample size per arm. One curve per dose. Shaded region = Monte Carlo 95% CI (n = 1000 replicates). Horizontal dashed line = 80% target."}
pow_path <- here("projects", "05_trial_simulation", "results", "power_curve.png")
if (file.exists(pow_path)) {
  knitr::include_graphics(pow_path)
} else {
  cat("Run 03_diagnostics.R to generate power curve.")
}
```

```{r oc-table}
oc_path <- here("projects", "05_trial_simulation", "results", "operating_characteristics.csv")
if (file.exists(oc_path)) {
  read.csv(oc_path) |>
    gt() |>
    tab_header(
      title = "Operating Characteristics",
      subtitle = "By dose and sample size; success = EC50 within ±30% of truth"
    ) |>
    fmt_percent(columns = c(Power, TypeIError), decimals = 1) |>
    fmt_number(columns = c(EC50_Bias, EC50_RMSE), decimals = 3) |>
    tab_style(
      style = cell_fill(color = "#d4edda"),
      locations = cells_body(rows = Power >= 0.80)
    ) |>
    tab_footnote("Highlighted rows achieve ≥ 80% power. TypeIError estimated
                  from placebo-vs-placebo replicates under the null (Emax = 0).")
} else {
  cat("Run 03_diagnostics.R to generate operating characteristics table.")
}
```

```{r ec50-precision, fig.width=9, fig.height=5, fig.cap="Distribution of EC50 estimates across 1000 replicates, by sample size and dose. Box = IQR; whiskers = 5th–95th percentile. Dashed line = true EC50. Wider spread at low n and low dose reflects identifiability limits."}
prec_path <- here("projects", "05_trial_simulation", "results", "ec50_precision.png")
if (file.exists(prec_path)) {
  knitr::include_graphics(prec_path)
} else {
  cat("Run 03_diagnostics.R.")
}
```

## Interpretation

```{r recommendation, results='asis'}
rec_path <- here("projects", "05_trial_simulation", "results", "sample_size_recommendation.txt")
if (file.exists(rec_path)) {
  cat(readLines(rec_path), sep = "\n")
} else {
  cat("**Sample size recommendation:** Generated after running 02_fit_model.R and 03_diagnostics.R.")
}
```

## Limitations

- Replication uses a Monte Carlo estimator; power estimates have simulation
  error of approximately ±1.5% at n = 1000 replicates (binomial SE).
- The analysis model (non-linear least squares on arm means) ignores IIV;
  a mixed-effects PK/PD model in the analysis step would give tighter EC₅₀
  estimates, so these power estimates are conservative.
- A fixed true EC₅₀ (2.5 mg/L) is used; uncertainty in EC₅₀ from the
  Phase 1 study would widen the precision requirements.
- Type I error is evaluated at the dose-level (each dose vs placebo); a
  family-wise error rate correction for multiple comparisons is not applied.
- Dropout and protocol deviations are not simulated.
