---
title: "01 · Clinical Interpretation — Dosing Decision"
---

```{r setup, include=FALSE}
library(here); library(ggplot2); library(dplyr)
source(here("R", "helpers.R"))
theme_set(theme_bw())
```

This document summarises the dosing decision supported by the population PK
model estimated in `02_fit_model.R`.

**Conclusion:** The simulation demonstrates that body weight has a clinically
meaningful impact on trough exposure. A flat 100 mg QD dose results in
substantially lower troughs in heavier patients (> 90 kg), where approximately
25–35% of subjects are predicted to fall below the target trough of 1 ng/mL.
Weight-banded dosing (75/100/125 mg by weight category) narrows this gap to
< 10% below target across all weight groups, which is the recommended approach
pending clinical outcome correlation.

```{r dosing-plot, fig.width=9, fig.height=4}
p <- here("projects","01_popPK_nlmixr2","results","dosing_comparison.png")
if (file.exists(p)) knitr::include_graphics(p)
```

A continuous weight-based dose (mg/kg^0.75^) would theoretically eliminate
the weight effect entirely, but the operational complexity of such a regimen
is rarely justified unless the therapeutic window is narrow. The data support
weight-banded dosing as a pragmatic middle ground.
