---
title: "Project 01 — Population PK: Flat vs Weight-Based Dosing"
subtitle: "2-compartment oral model with allometric scaling · nlmixr2 SAEM"
date: today
---

```{r setup, include=FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(gt)
library(kableExtra)
source(here("R", "helpers.R"))
theme_set(theme_bw())
```

> **Simulated data.** All concentrations in this report were generated from a
> known pharmacokinetic model with documented true parameters. No real patient
> data are used.

## Background

Weight is the most commonly investigated covariate in population PK models
because body size influences apparent volume of distribution and renal/hepatic
clearance. Allometric theory predicts clearance scales with body weight (BW)
raised to the 0.75 power and volume scales linearly. Whether this relationship
is clinically meaningful enough to require weight-based dosing — rather than a
flat dose — depends on the magnitude of the exposure difference across the
observed weight range and the therapeutic window.

## Question

Does a flat 100 mg once-daily (QD) oral dose maintain target trough
concentration (C~trough~ > 1 ng/mL) across the body weight range 40–120 kg,
or is weight-based dosing (or weight-banded dosing) needed?

## Data

```{r load-data}
pk <- read.csv(here("projects", "01_popPK_nlmixr2", "data", "pk_data.csv"))
cat(sprintf("Subjects: %d | Observations: %d | Weight range: %.0f–%.0f kg\n",
    length(unique(pk$ID)), sum(pk$EVID == 0),
    min(pk$BW[pk$EVID==0]), max(pk$BW[pk$EVID==0])))
```

```{r data-summary-table}
pk |>
  filter(EVID == 0) |>
  group_by(ID) |>
  summarise(BW = first(BW), AGE = first(AGE), SEX = first(SEX),
            DOSE = first(DOSE), n_obs = n(), .groups = "drop") |>
  summarise(
    `N subjects` = n(),
    `BW mean (SD)` = sprintf("%.1f (%.1f)", mean(BW), sd(BW)),
    `Age mean (SD)` = sprintf("%.1f (%.1f)", mean(AGE), sd(AGE)),
    `% female` = sprintf("%.0f%%", 100*mean(SEX==1)),
    `Obs per subject (median)` = median(n_obs)
  ) |>
  gt() |>
  tab_header(title = "Simulated Study Demographics") |>
  tab_footnote("All values are from a synthetic dataset. True PK parameters
               are documented in 01_simulate_data.R.")
```

## Model

**Structural model:** two-compartment pharmacokinetics with first-order oral
absorption from a depot compartment.

$$
\frac{dA_{dep}}{dt} = -K_a \cdot A_{dep}
$$
$$
\frac{dA_{c}}{dt} = K_a \cdot A_{dep} - \frac{CL}{V_c} A_c - \frac{Q}{V_c} A_c + \frac{Q}{V_p} A_p
$$
$$
\frac{dA_{p}}{dt} = \frac{Q}{V_c} A_c - \frac{Q}{V_p} A_p
$$
$$
C = A_c / V_c
$$

**Allometric covariate model:**

$$
CL_i = CL_{ref} \cdot \left(\frac{BW_i}{70}\right)^{0.75} \cdot e^{\eta_{CL,i}}
$$
$$
V_{c,i} = V_{c,ref} \cdot \left(\frac{BW_i}{70}\right)^{1.0} \cdot e^{\eta_{V,i}}
$$

**IIV:** log-normal on CL, V~c~, and K~a~.
**RUV:** proportional + additive.
**Estimation:** nlmixr2 SAEM (n.burn = 200, n.em = 300).

## Results

```{r parameter-table}
if (file.exists(here("projects", "01_popPK_nlmixr2", "results", "parameter_table.csv"))) {
  par_tbl <- read.csv(here("projects", "01_popPK_nlmixr2", "results", "parameter_table.csv"))
  par_tbl |>
    gt() |>
    tab_header(title = "Population PK Parameter Estimates",
               subtitle = "2-compartment oral model with allometric weight") |>
    fmt_number(columns = where(is.numeric), decimals = 3) |>
    tab_footnote("RSE = relative standard error from Fisher information matrix.
                  Bootstrap 95% CI from n = 200 case-resampling replicates.")
} else {
  cat("Run 02_fit_model.R to generate parameter estimates.")
}
```

## Diagnostics

```{r gof-panel, fig.width=10, fig.height=9, fig.cap="GOF panel: DV vs PRED (upper left), DV vs IPRED (upper right), CWRES vs TIME (lower left), CWRES vs PRED (lower right). Reference lines at 0 (solid) and ±1.96 (dashed)."}
gof_path <- here("projects", "01_popPK_nlmixr2", "results", "gof_panel.png")
if (file.exists(gof_path)) {
  knitr::include_graphics(gof_path)
} else {
  cat("Run 03_diagnostics.R to generate GOF plots.")
}
```

```{r vpc-plot, fig.width=10, fig.height=5, fig.cap="Prediction-corrected VPC stratified by body weight tertile (low: <60 kg, mid: 60–85 kg, high: >85 kg). Ribbons show 95% CI of simulated 5th/50th/95th percentiles (n = 1000 simulations). Black dashed lines are observed percentiles."}
vpc_path <- here("projects", "01_popPK_nlmixr2", "results", "pcvpc_by_weight.png")
if (file.exists(vpc_path)) {
  knitr::include_graphics(vpc_path)
} else {
  cat("Run 03_diagnostics.R to generate VPC.")
}
```

```{r shrinkage}
shrink_path <- here("projects", "01_popPK_nlmixr2", "results", "shrinkage_table.rds")
if (file.exists(shrink_path)) {
  readRDS(shrink_path)
} else {
  cat("Run 03_diagnostics.R to generate shrinkage table.")
}
```

```{r bootstrap-ci}
boot_path <- here("projects", "01_popPK_nlmixr2", "results", "bootstrap_summary.rds")
if (file.exists(boot_path)) {
  readRDS(boot_path)
} else {
  cat("Run 03_diagnostics.R to generate bootstrap CI table.")
}
```

## Interpretation

```{r dosing-simulation, fig.width=9, fig.height=5, fig.cap="Simulated steady-state trough concentration distributions by dosing strategy and body weight group. Horizontal dashed line = target trough (1 ng/mL). Numbers above bars show % of subjects below target."}
dose_path <- here("projects", "01_popPK_nlmixr2", "results", "dosing_comparison.png")
if (file.exists(dose_path)) {
  knitr::include_graphics(dose_path)
} else {
  cat("Run 03_diagnostics.R to generate dosing simulation plot.")
}
```

```{r decision, results='asis'}
dec_path <- here("projects", "01_popPK_nlmixr2", "results", "dosing_decision.txt")
if (file.exists(dec_path)) {
  cat(readLines(dec_path), sep = "\n")
} else {
  cat("**Dosing decision:** Generated after running 02_fit_model.R and 03_diagnostics.R.")
}
```

## Limitations

- Dataset is synthetic; true model structure is known to match the estimation
  model, which is not typical in practice.
- A fixed allometric exponent (0.75 for CL) was assumed; the data-estimated
  exponent is reported in the parameter table as a sensitivity check.
- The target trough (1 ng/mL) is illustrative and not linked to a clinical
  endpoint in this portfolio.
- Bootstrap was run with n = 200 replicates; for regulatory submissions
  n ≥ 500 is recommended.
- Only body weight was evaluated as a covariate; age, renal function, and
  sex may also be relevant.
